---
title: "Project 4 Code"
author: "Charley Ferrari"
date: "July 11, 2016"
output: pdf_document
---

```{r warning=FALSE, message=FALSE}

library(plyr)
library(dplyr)
library(reshape2)

library(plotly)
library(Matrix)

library(recommenderlab)

setwd('/Users/Charley/Downloads/cuny/IS 643 Recommender Systems/Week 4')

```

## Adding Context to a Recommendation Engine

This dataset comes from a collaborative bookmark tagging platform called del.icio.ous. In this platform, users are given the ability to store bookmarks and tag them. Our main data matrix, user_taggedbookmarks, includes a collection of timestamped tags. My business case will be to recommend bookmarks to users, based on information given by their choices in bookmarks and tags. For this system, I will demonstrate the multiple dimensions I have by creating two separate measures for user/user similarity: using tags and using bookmarks. In reality, this multiplies the choices of similarity I can use. For example, I could use an item/item approach and use either tags or users to measure similarity. I'm choosing to stick with user/user because it makes the most sense for a collaborative recommendation model (the tag/bookmark comparison makes more sense for a content-based approach.)

I'm also going to add time as a context. I'll calculate the age of a bookmark by its first appearance in the database. I'll take the time of the recommendation into account, and weight a bookmark's recommendation by how old it is (I will also use this method to remove bookmarks that haven't been added to the system yet.)

First I'll bring in my data and calculate my bookmark ages:

```{r}
user_taggedbookmarks_timestamps <- read.delim('user_taggedbookmarks-timestamps.dat',
                                              sep='\t', header=TRUE)

user_taggedbookmarks_timestamps$normTime <- 
  user_taggedbookmarks_timestamps$timestamp -
  min(user_taggedbookmarks_timestamps$timestamp)

bookmarkAge <- user_taggedbookmarks_timestamps %>%
  dplyr::group_by(bookmarkID) %>%
  summarize(Appearance = min(timestamp))

user_taggedbookmarks_timestamps <- merge(user_taggedbookmarks_timestamps,
                                         bookmarkAge, by='bookmarkID')

```

These actions are binary, but when we're looking at only two variables, users by bookmarks or users by tags, we can get multiple counts (users can add multiple tags to a bookmark for example.) I'm going to treat this differently depending on whether we're looking at users and bookmarks or users and tags. 

For users and bookmarks, increasing the number of tags and bookmarks does not increase the affinity of that user to a bookmark. The user has already expressed interest in this bookmark, and is simply adding additional classification to it. By contrast, adding multiple bookmarks to a tag does increase a user's affinity for that tag. It proves they more interested in that topic matter, and are reading more links about it.

So when I'm calculating my user/bookmark and user/tag matrix, the user/bookmark matrix will include only 1's, while the user/tag matrix will include weights for each tag.

```{r}

userbookmarktagFAC <- user_taggedbookmarks_timestamps %>%
  transform(bookmarkID = factor(bookmarkID), userID = factor(userID), 
            tagID = factor(tagID))

userbookmarkBIN <- userbookmarktagFAC %>%
  distinct(userID, bookmarkID) %>% select(userID, bookmarkID)

usersBYbookmarks <- sparseMatrix(as.integer(factor(userbookmarkBIN$bookmarkID)),
               as.integer(factor(userbookmarkBIN$userID)),
               x=1)

usersBYtags <- sparseMatrix(as.integer(factor(userbookmarktagFAC$tagID)),
               as.integer(factor(userbookmarktagFAC$userID)),
               x=1)

```

Next, I'll calculate two user similarity matrices:

```{r}

sumsquares <- function(x){
  return(sqrt(sum(x^2)))
}

userSimUB <- aaply(usersBYbookmarks, 1, sumsquares)
userMultUB <- usersBYbookmarks %*% t(usersBYbookmarks)
userSimOuterUB <- outer(userSimUB, userSimUB, '*')
useruserUB <- userMultUB / userSimOuterUB

userSimUT <- aaply(usersBYtags, 1, sumsquares)
userMultUT <- usersBYtags %*% t(usersBYtags)
userSimOuterUT <- outer(userSimUT, userSimUT, '*')
useruserUT <- userMultUT / userSimOuterUT

```



















```{r eval=FALSE}

bookmark_tags <- read.delim('bookmark_tags.dat', header=TRUE)

bookmarks <- read.delim('bookmarks.dat', sep='\t', header=TRUE)

tags <- read.delim('tags.dat', sep='\t', header=TRUE)







plot_ly(user_taggedbookmarks_timestamps, x=normTime, y=userID, type='histogram')

qplot(user_taggedbookmarks_timestamps$Date)





user_taggedbookmarks_timestamps$Date <- 
  as.POSIXct(user_taggedbookmarks_timestamps$timestamp, origin="1970-01-01")

sd(user_taggedbookmarks_timestamps$timestamp)

```

## Adding context to a recommendation engine



I tried to apply the models put forth in chapter 10 of Statistical Methods for Recommender Systems, but believe that a simpler method will be possible for this dataset, and will only be taking the core concept of thinking of this recommendation system as a three dimensional problem.

Rather than just dealing with users and items, we will consider tags to be our third dimension. In two dimensional versions of this problem, we would compute the cosine similarity between users and/or items. In this case, we are computing a version of the cosine similarity (based on the Frobenius norm) between vectors, and can judge similarities between users, bookmarks, or tags.

The objective of this recommendation engine will be to compute the probability of a user tagging a bookmark with a particular tag. By computing distance in this way, we should be ensuring that both the tag and the bookmark will be relevent to the user.

I couldn't find any great solutions to dealing with sparse three dimensional arrays, but because we are dealing with binary data (0's for no interaction or 1 for a successful tag), the calculations should be relatively straight forward. The similarity measure I'm using is shown below:

$$ \frac{\sum^{n_i}_{i=1} \sum^{n_j}_{j=1} A_{ij} B_{ij}}{\sqrt{\sum^{n_i}_{i=1} \sum^{n_j}_{j=1} A_{ij}} \sqrt{\sum^{n_i}_{i=1} \sum^{n_j}_{j=1} B_{ij}}} $$

In our problem, for computing two users similarities, this simplifies. Lets consider a to be the number of tags by user A, b to be the number of tags by user B, and s to be the number of tag-bookmark pairs that are shared. We end up with

$$ \frac{s}{a \times b} $$

So, I will write my similarity functions to take advantage of this directly from the data frame. Here is what this function would look like for a user:

```{r}

userSimCalc <- function(u1, u2){
  slice <- user_taggedbookmarks_timestamps %>%
    filter(userID %in% c(u1, u2))
  numdistinct <- nrow(
    slice %>% distinct(bookmarkID, tagID)
  )
  s <- nrow(slice) - numdistinct
  u1nr <- nrow(
    slice %>% filter(userID == u1)
  )
  u2nr <- nrow(
    slice %>% filter(userID == u2)
  )
  return(s / (u1nr*u2nr))
}

plot_ly(user_taggedbookmarks_timestamps, x=timestamp, y=userID, type=)

```
